---
title: "Coordinator Performance Task"
output-file: "2024-07-26-Steven-Primeaux-Performance Task"
format:
  html:
    embed-resources: true
    code-fold: true
    toc: true
  docx: default
execute: 
  message: false
  warning: false
---

## Introduction

This data notebook contains all of the code needed for the calculations and 
visualizations, as well as the text of the email. Section 1, Project Setup, 
contains some introductory code chunks needed for the analysis, while Section 2,
Email, contains the text of the email itself, including visualizations.

## Section 1: Project Setup

Load package dependencies.

```{r}
library(tidyverse)
library(readxl)
library(writexl)
library(janitor)
library(scales)
library(knitr)
```

Create utility variables for plotting, including:

* The number of full months assumed to have already elapsed this fiscal year.
* An array of months for time-series axes.
* An array of keys to indicate solid or dashed lines (actual versus projected).

```{r}
n_months_elapsed <- 7

months_vector <- seq.Date(
  as_date("2023-10-01"), as_date("2024-09-01"),
  by = "month"
)

linetypes_vector <- factor(c(1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2))
```

Read in raw data.

```{r}
data_raw <-
  read_excel("data.xlsx", range = "A18:AD6740") |>
  clean_names()
```

Create new fields for analysis.

```{r}
data <-
  data_raw |>
  mutate(budgeted = round(total_budget, 0)) |>
  mutate(committed = round(expenditure + obligation + commitment, 0)) |>
  mutate(available = round(available_budget, 0)) |>
  mutate(
    budgeted_month    = round(budgeted / 12, 0),
    committed_month   = round(committed / n_months_elapsed, 0),
    budgeted_current  = budgeted_month * n_months_elapsed,
    committed_current = committed_month * n_months_elapsed,
    variance_abs      = committed_current - budgeted_current,
    variance_prop     = variance_abs / budgeted_current,
    available_current = budgeted_current - committed_current
  ) |>
  mutate(across(c(cost_center), as.character)) |>
  rename(
    account_3 = account_category_description_parent_level_3,
    account_1 = account_group_parent_level_1_description,
    account_0 = account_description
  )
```

## Section 2: Email

Dear Ms. Smith:

Thanks so much for your email. I'll be happy to answer those questions, and I've
also included some charts that I found helpful in thinking about these questions
myself.

### Are we on track for this point of the school year?

```{r}
accounts_3_cum <-
  data |>
  group_by(account_3) |>
  summarize(
    budgeted_month = sum(budgeted_month),
    committed_month = sum(committed_month)
  ) |>
  mutate(n = 12) |>
  uncount(n) |>
  group_by(account_3) |>
  mutate(
    Budgeted  = cumsum(budgeted_month),
    Committed = cumsum(committed_month)
  ) |>
  mutate(month_fiscal = months_vector, linetype = linetypes_vector) |>
  select(account_3, month_fiscal, Budgeted, Committed, linetype) |>
  pivot_longer(
    cols = c(Budgeted, Committed), names_to = "var", values_to = "val"
  )
```

```{r}
data |>
  group_by(account_3) |>
  summarize(
    budgeted = sum(budgeted),
    committed = sum(committed),
    budgeted_current = sum(budgeted_current),
    committed_current = sum(committed_current)
  ) |>
  mutate(
    variance_abs = committed_current - budgeted_current,
    variance_prop = variance_abs / budgeted_current
  ) |>
  mutate(across(account_3, \(x) str_remove(x, " SERVICES"))) |>
  arrange(desc(budgeted)) |>
  ungroup() |>
  kable()
```

#### Personnel

```{r}
personnel <-
  accounts_3_cum |>
  filter(account_3 == "PERSONNEL SERVICES")

personnel |>
  ggplot(aes(month_fiscal, val, color = var)) +
  geom_line(aes(linetype = linetype)) +
  geom_line(
    data = filter(personnel, month_fiscal <= "2024-04-01"), linetype = 1
  ) +
  scale_y_continuous(labels = label_currency(scale_cut = cut_long_scale())) +
  scale_linetype(guide = NULL) +
  ggtitle("Personnel Services") +
  xlab("Month") +
  ylab("Amount") +
  labs(color = NULL)
```

Starting with our personnel budget of $1.186 billion, through the first seven 
months of the fiscal year (through April), we'd expect to have committed around 
\$692 million by this point. At \$590 million, we're around $101 million under 
budget, or 15 percent.

#### Non-Personnel

```{r}
nonpersonnel <-
  accounts_3_cum |>
  filter(account_3 == "NON-PERSONNEL SERVICES")

nonpersonnel |>
  ggplot(aes(month_fiscal, val, color = var)) +
  geom_line(aes(linetype = linetype)) +
  geom_line(
    data = filter(nonpersonnel, month_fiscal <= "2024-04-01"), linetype = 1
  ) +
  scale_y_continuous(labels = label_dollar(scale_cut = cut_long_scale())) +
  scale_linetype(guide = NULL) +
  ggtitle("Non-Personnel Services") +
  xlab("Month") +
  ylab("Amount") +
  labs(color = NULL)
```

Looking now at our non-personnel budget of $244 million, through the first seven 
months of the fiscal year (through April), we'd expect to have committed around 
\$142 million by this point. At \$123 million actual, we're around $19 million 
under budget, or 13 percent.

### Are there any budget concerns? Where are we over-spending?

#### Accounts

```{r}
accounts_0 <-
  data |>
  group_by(account_3, account_0) |>
  summarize(
    budgeted_current = sum(budgeted_current),
    committed_current = sum(committed_current)
  ) |>
  mutate(
    variance_abs = committed_current - budgeted_current,
    variance_prop = variance_abs / budgeted_current
  ) |>
  arrange(desc(account_3), desc(budgeted_current)) |>
  ungroup()
```

```{r}
accounts_0 |>
  filter(variance_prop > 0) |>
  ggplot(aes(variance_prop, fct_reorder(account_0, variance_prop))) +
  geom_col() +
  scale_x_continuous(labels = label_percent()) +
  ggtitle("Budget Variance - Percent", "Accounts") +
  xlab("Percent") +
  ylab("Category")
```

Looking first at the account level, as a percent of budgeted amount, auto 
maintenance and repairs, security services, and IT software maintenance appear 
to be most over budget for this point in the year.

```{r}
accounts_0 |>
  filter(variance_abs > 0) |>
  ggplot(aes(variance_abs, fct_reorder(account_0, variance_abs))) +
  geom_col() +
  scale_x_continuous(labels = label_currency(scale_cut = cut_long_scale())) +
  ggtitle("Budget Variance - Dollars", "Accounts") +
  xlab("Dollars") +
  ylab("Category")
```

By absolute dollar value, land and structure rentals, gas, and security 
services appear to be most over budget.

#### Cost Centers

```{r}
costcenters <-
  data |>
  group_by(account_3, cost_center) |>
  summarize(
    budgeted_current = sum(budgeted_current),
    committed_current = sum(committed_current)
  ) |>
  mutate(
    variance_abs = committed_current - budgeted_current,
    variance_prop = variance_abs / budgeted_current
  ) |>
  arrange(desc(account_3), desc(budgeted_current)) |>
  ungroup()
```

```{r}
costcenters |>
  filter(variance_prop > 0) |>
  ggplot(aes(variance_prop, fct_reorder(cost_center, variance_prop))) +
  geom_col() +
  scale_x_continuous(labels = label_percent()) +
  ggtitle("Budget Variance - Percent", "Cost Centers") +
  xlab("Percent") +
  ylab("Category")
```

Looking now at the cost center level, as a percent of budgeted amount, 40122, 
40036, and 40048 appear to be most over budget for this point in the year.

```{r}
costcenters |>
  filter(variance_abs > 0) |>
  ggplot(aes(variance_abs, fct_reorder(cost_center, variance_abs))) +
  geom_col() +
  scale_x_continuous(labels = label_currency(scale_cut = cut_long_scale())) +
  ggtitle("Budget Variance - Dollars", "Cost Centers") +
  xlab("Dollars") +
  ylab("Category")
```

By absolute dollar value, 40122 and 40076 are currently over by more than 
\$75,000, with 40036 additionally being over budget by more than \$50,000.

### Are there any risks of underspending over $100,000 for any cost center or account?

Note that responses in this section are focusing on non-personnel expenses, 
which may be more relevant for the June 7 procurement deadline.

#### Accounts

```{r}
#| eval: false

accounts_0 |>
  filter(account_3 == "NON-PERSONNEL SERVICES") |>
  filter(variance_abs < -100000) |>
  arrange(variance_abs)
```

At the account level, 25 accounts are potentially underspending at this point in
the year by more than $100,000. 

```{r}
accounts_0 |>
  filter(account_3 == "NON-PERSONNEL SERVICES") |>
  filter(variance_abs < -100000) |>
  ggplot(
    aes(variance_abs, fct_reorder(account_0, variance_abs, .desc = TRUE))
  ) +
  geom_col() +
  scale_x_continuous(labels = label_currency(scale_cut = cut_long_scale())) +
  ggtitle("Largest Potential Underspend") +
  xlab("Dollars") +
  ylab("Category")
```

For the account level, at this point in the year professional service fees and 
contracting are potentially underspending by more than $4 million, while three 
more accounts are potentially underspending by more than $1 million.

#### Cost Centers

```{r}
#| eval: false

costcenters |>
  filter(account_3 == "NON-PERSONNEL SERVICES") |>
  filter(variance_abs < -100000) |>
  arrange(variance_abs) |>
  kable()
```

At this point in the year, 63 cost centers are potentially underspending by more
then $100,000.

```{r}
costcenters |>
  filter(account_3 == "NON-PERSONNEL SERVICES") |>
  filter(variance_abs < -100000) |>
  slice_min(order_by = variance_abs, n = 10) |>
  ggplot(
    aes(variance_abs, fct_reorder(cost_center, variance_abs, .desc = TRUE))
  ) +
  geom_col() +
  scale_x_continuous(labels = label_currency(scale_cut = cut_long_scale())) +
  ggtitle("Largest Potential Underspend") +
  xlab("Dollars") +
  ylab("Category")
```

At this point in the year, cost centers 40086 and 40078 were both potentially 
underspending by more than $1 million.

### Immediate Pressures

#### Chromebooks

```{r}
data |>
  filter(account_0 == "IT HARDWARE ACQUISITIONS") |>
  group_by(fund, account, fund_description, account_0) |>
  summarize(
    budgeted = sum(budgeted),
    committed = sum(committed),
    available = sum(available)
  ) |>
  arrange(desc(available)) |>
  kable()
```

If we look at the full-year budgeted amount for IT hardware acquisitions, 
$250,000 for the Chromebooks would potentially be available under local funds, 
Medicaid Public Provider Recovery Grant, and At-Risk Concentration individually.

```{r}
data |>
  filter(account_0 == "IT HARDWARE ACQUISITIONS") |>
  group_by(fund, account, fund_description, account_0) |>
  summarize(
    budgeted_current = sum(budgeted_current),
    committed_current = sum(committed_current),
    available_current = sum(available_current)
  ) |>
  arrange(desc(available_current)) |>
  kable()
```

However, if we instead calculate a year-to-date budgeted amount (assuming seven 
months have elapsed in the fiscal year), the full $250,000 would only be 
available under local funds, although part of the amount could still be moved 
from Medicaid Public Provider Recovery Grant and At-Risk Concentration.

Another consideration is that At-Risk Concentration and Medicaid funds could be 
restricted, so in this case local funds probably make more sense.

In conclusion, budget code GAO-1010-1010001-7171008 would be my recommendation.

#### Field Trip

```{r}
data |>
  filter(account_0 %in% c("PROF SERVICE FEES & CONTR", "TRAVEL - LOCAL")) |>
  filter(appropriated_fund_description == "LOCAL FUND") |>
  filter(fund_description != "AT-RISK CONCENTRATION") |>
  group_by(fund, account, fund_description, account_0) |>
  summarize(
    budgeted_current = sum(budgeted_current),
    committed_current = sum(committed_current),
    available_current = sum(available_current)
  ) |>
  arrange(desc(available_current)) |>
  kable()
```

After filtering for the professional services and local travel accounts, then 
restricting our search to local fund appropriations and excluding At-Risk 
Concentration funds, it does appear that we can shift some resources for the 
field trip costs.

For the $8,000 needed for local buses, I would recommend budget code 
GAO-1010-1010001-7131002, and for the $28,324 needed under professional services
I would recommend budget code GAO-1010-1010001-7131009.

```{r}
write_xlsx(
  list(
    "Data" = data,
    "Top Level Monthly" = accounts_3_cum,
    "Accounts YTD" = accounts_0,
    "Cost Centers YTD" = costcenters
  ),
  path = "2024-07-26-Steven-Primeaux-Performance Task.xlsx"
)
```

I've attached my working files for reference. Please let me know if you have any
other questions, I'll be happy to assist.

--Steven

```{r}
Sys.time()
```
